### 2009-04-18 response-to-reviews2.txt

> ------------------------------------------------------------------------
>
> Reviewer #1 (Remarks for the Author):
>
> This new draft is vastly improved. (One comment I penned into the 
> margin at the end of the introduction is "lightyears better"). The 
> abstract and introduction now clearly state the problem and its 
> background. The methods are now clear enough up to about page 10 of 
> the manuscript. After that, however, the manuscript breaks down. It is 
> still too vague for me.
>
> It's now much clearer that the manuscript promises two main results: a 
> "transducer composition algorithm" for building a multisequence SCFG 
> from pairwise transducers, given a tree; and a dynamic programming 
> algorithm for calculating maximum likelihood parses under this 
> grammar, given N input sequences. The results consist of implementing 
> the DP algorithm for a particular RNA evolutionary model (called 
> "TKFST") and trying it out on a couple of small toy examples of three 
> short RNA sequences (nanos TCEs, and tRNAs). Though one can get the 
> rough gist of how the authors' approach works, many important details 
> necessary for an acceptable level of understanding and reproducibility 
> still seem to be missing.
>
> The problem statement (p.5) is ok but still needs some clarification, 
> and perhaps some discussion from the perspective of where the 
> rate-limiting problems in "paleogenetic reconstruction" really lie. 
> What exactly is given and what exactly is going to be inferred? Does 
> "Given a phylogenetic tree" mean that both tree topology and its 
> branch lengths are assumed known? (The examples in Figures 1 and 2 
> seem to indicate so.) Are individual secondary structures assumed to 
> be known for the observed sequences? (Again, Figures 1 and 2 suggest 
> so.) And from this are we inferring the alignment of the observed 
> sequences, and the alignment, secondary structure, and sequence of the 
> unobserved ancestors? Aside from clarification, one reason that it 
> would be good to nail this down is to bring out that the authors 
> apparently believe that it is reasonable to consider individual 
> secondary structures to be "known" in the input, whereas the alignment 
> is an "unknown" that must be inferred. To my mind,
> this would get the rate-limiting inferences backwards. If I were 
> really trying to reconstruct the evolution of an important RNA, I 
> would have a lovingly handcrafted multiple alignment already, much 
> better than any automated inference program would be likely to produce 
> (c.f. Gutell rRNA alignments, or catalytic intron alignments) as well 
> as a highly accurate consensus secondary structure, but individual 
> secondary structures (where they differed from consensus) would remain 
> in doubt, and would need to be inferred. Not that structural alignment 
> isn't of interest; just that it's not obvious that it's the most 
> important task if paleogenetic reconstruction is the real
> motivation.

Thank you for the detailed recommendations.  We now state clearly that
we take the phylogeny as given, although we also note in the
Discussion that in principle we could construct a basic sampler
over trees using our method (though we have not constructed or
evaluated such a sampler).

We have added more discussion of the potential rate-limiting steps and
corresponding constraints to the manuscript ("Overview").  As the
reviewer points out, perhaps the most realistic pieces of outside
information to have are a consensus structure for sequences and/or a
curated multiple alignment.  In principle these can all be
incorporated into our inference algorithms with the alignment
constraints used by e.g. QRNA and PFOLD, or the simultaneous fold and
alignment constraints used by e.g. Stemloc and CONSAN.

For this paper we used constraints on individual structures because
these are simple to implement.  We state this clearly in the revised
text and note that while experimentally-determined structures of
individual RNAs are rare, curated multiple alignments are
more-commonly available and can be used to constrain the algorithms
which we describe.

In principle we think that the reconstruction might benefit as well
from partial alignment constraints as from partial fold constraints;
that is, as well as being able to deviate slightly from the consensus
fold (as the reviewer suggests), the reconstruction algorithm would
probably need to deviate slightly from the supplied alignment as
well. (This will be necessary because not every evolutionarily related
basepair is likely to be perfectly aligned.) This sort of flexible
fold and alignment constraint is, of course, implemented in CONSAN and
Stemloc. However, we note that the Pair SCFGs used in both those
algorithms were first implemented using much simpler constraints:
specifically, the fold-only constraints of (Holmes 2002), and the
alignment-only constraints of (Rivas & Eddy 2001).

We do hope that the sophistication of the constraint schemes used with
these "tree transducer" models will increase over time. Eventually we
hope to develop alignment constraints, simultaneous alignment-and-fold
constraints, and other constraints (and/or sampling schemes) to enable
efficient analysis of multiple phylogenetically-related sequences. The
present work represents a first attempt at using these sorts of models
for more than two sequences, but we hope it will not be the last
attempt, and we acknowledge room for improvement.

> It is still not clear eenough what the authors' model is, though it is 
> supposed to be a main result. Page 7 states "a prescriptive formalism 
> for transducers is given in Text S1". Relegating the deliverables of a 
> manuscript to supplementary material is not an appropriate use of 
> supplementary material; a reader should not have to refer to a second 
> manuscript to understand the main points. The authors need to find a 
> way to explain their main results in the actual manuscript. There is 
> still much vagueness and redundancy in this draft, which could be 
> removed and replaced with more relevant explanation. For example, in 
> this same section where results are relegated to supplementary 
> material, there's a whole redundant paragraph (p.8) on how conditional 
> models represent directed edges; but this concept has already been 
> explained sufficiently. This is followed (bottom p. 8, "Consider a 
> phylogenetic tree...") by yet another statement of the same concept of 
> using conditional models, still in
> terms too vague and conceptual to enable reproduction of the work.

Thank you for pointing this out.  We have removed the reference to the
supplementary material in this section and have attempted to make the
main text a self-contained description of the composition algorithm.
We have attempted to remove redundant text, including the two
paragraphs which the reviewer notes.

> It is still not clear enough what the authors mean by a "composition 
> algorithm". The description of the generative model on pp 9-10 seems 
> redundant with the "composition algorithm" on pp 11-12; pp 11-12 don't 
> seem to add anything that I didn't already understand from pp 9-10, 
> and is still not precise and concrete enough for me to really 
> understand the actual model in play here. My understanding is that a 
> "composition algorithm" is just how one chains conditional models 
> together to calculate the desired joint probability of the observed 
> data. Conceptually, this chaining seems obvious enough to me when a 
> phylogenetic tree is given, and the use of so much space to describe 
> the concept is puzzling; but perhaps I'm missing something. In any 
> case, what the reader really wants is a precise explanation of the 
> exact model in use here, not a lot of text about the well-understood 
> concept of conditional likelihoods on branches of a tree.

We have reorganized the presentation of the "composition algorithm" to
attempt to clarify both what it is and why it is important.  The
reviewer is completely correct that it is, indeed, a prescription
for chaining together conditional models.  However, while the concept
of so chaining together models is intuitive, in practice it can be a
daunting task for SCFGs thanks to the presence of bifurcations.  For
example, while both the model generating a prior distribution over ancestral 
sequences (the "singlet transducer," with 7 states) and the model describing
evolution along a single branch (the "branch transducer," with 21 states)
are simple and compact, the three-taxon ensemble model has 230 states
and 1,789 transitions.

In summary, while we completely agree that the model construction procedure
is no more than a formalization of an intuitive concept, we believe
that it is nonetheless substantial, due to the sheer complexity
of the ensemble models so created.  The "TKFST" model which we use
here is relatively simple; more realistic models will give rise to even
more complex ensemble models.

> Perhaps most glaringly, the section entitled "dynamic programming 
> inference algorithms" never actually states an algorithm! It should. 
> The "loopy DP" algorithm is claimed in the abstract (and in the 
> authors' response to reviews) to be a key result of the paper, but it 
> is only referred to ("See Text S1 for more [sic] details of this 
> algorithm"). Again this is a misuse of supplementary material. A paper 
> is not supposed to be a mere advertisement of cool things we're 
> working on, and cooler things we're thinking about working on. Too 
> much of this manuscript still reads this way (though it's greatly 
> improved from the previous version). A paper should deliver crisp, 
> hard-nosed, reproducible results to a reader. In the case of an 
> algorithms paper, the reader should be able to understand and 
> implement the key ideas from a careful and complete reading of the 
> paper. Generally speaking, the authors are still spending way too much 
> time talking about ideas for their future, and too little
> teaching the reader something concrete in the present. Relegating 
> important stuff to supplementary material while spending so much room 
> on vague conceptual stuff in the text is aggravating (I would be 
> sympathetic if the manuscript were closely on point, and stuff was 
> being moved to supp material solely to keep length down and maintain 
> clarity of the essential results description). Supplementary material 
> should be used for stuff that doesn't fit (large data sets, source 
> code), is better distributed in electronic form (video, formatted 
> data) or is truly parenthetical to the main results (details that the 
> reader doesn't need to see to understand the paper). Only in a last 
> resort (in space-limited journals, which does not include PLoS 
> Computational Biology) should it contain any actual results of the paper.

Thank you for the feedback; we have attempted to revise the paper
accordingly such that it is completely self-contained and
reproducible.  We have given complete specifications of the CYK,
Inside, Outside, and CYK traceback algorithms for three sequences,
whether completely unconstrained, or constrained by a pre-determined
set of folds for each sequence ("fold envelopes").  They are written
in pseudocode which corresponds closely to our actual implementation
in the program (INDIEGRAM) which we describe.

We have entirely replaced the "loopy DP" algorithm with an algorithm
for exactly eliminating null cycles from general SCFGs.  Per the
reviewer's recommendation, the full description of this algorithm is
described in the main text ("Eliminating null cycles in SCFGs") of the
revised manuscript.  This new algorithm avoids the increased time
complexity associated with the iterative inference of "loopy DP."

> It's possible that I would come to adequately understand these 
> algorithms if I spent a lot more time thinking about them and studying 
> the "Supplementary" material. As always when refereeing, I feel I need 
> to allow for the possibility that I'm slower than the average reader. 
> So I'd actually be willing to forgo much of the above criticism if I 
> could skip ahead to the application and results and be impressed with 
> the utility of the approach, so that I'd feel motivated to go back and 
> work even harder to understand it. But the figures and results here 
> really do not inspire much confidence either.
>
> The key results figure is Figure 2, which shows the "reconstruction" 
> of "ancestral" sequences for three nanos TCEs and three tRNAs on star 
> topologies. The problems I have with this figure are manyfold, including:
>
> - In both cases the "ancestor" is in fact related by negligible branch 
> length to one of the observed sequences. But of course this 
> immediately makes the inference problem trivial, completely 
> uninteresting, and definitively *not* "ancient reconstruction". In 
> this situation, one can simply infer that the "ancestral" sequence is 
> identical to this closely related observed sequence. As noted in the 
> figure legend, that's almost what happens in the tRNA example (the 
> "ancestor" is nearly identical to AC008670.6). The figure legend fails 
> to note it also happens with the nanos TCEs, where the "ancestor" is 
> in fact completely identical to M72421.1. Even for a toy problem, come 
> on; this is way too unchallenging and toylike.
>
> - Even so, it looks like the inference algorithm is actually screwing 
> up the inference of the ancestor, moving in the wrong direction from 
> simply assuming identity with the near-zero-branch-length related 
> sequence. How does the algorithm possibly infer an ancestral G in a 
> column where it sees two gaps and an A (3' most base of the first stem 
> loop of tRNA)? How does it infer an ancestral gap in a column where it 
> sees two G's and an A and no gaps (the multifurcated base between the 
> D arm and the anticodon arm)? The authors need to explain how these 
> could possibly be right; they seem like red flags that the algorithm 
> isn't working properly.
>
> - This isn't an experiment. Showing that an algorithm produces a 
> result without crashing doesn't say much about whether the result 
> should be relied upon. If the purpose of these algorithms is to 
> accurately reconstruct ancestral RNAs, the authors need to design and 
> perform experiments to evaluate the accuracy of their reconstructions, 
> even if (for computational complexity reasons) these experiments can 
> only be done on a small, noncomprehensive set of anecdotes. I would 
> even settle for the normal positive control "sanity check" experiment 
> of simulating data under exactly their model, then verifying that 
> their model can reconstruct the appropriate likelihoods and inferences.

We agree that these reconstructions were not very informative.
Unfortunately these were the only sequences that we could find which
were short enough to be tractable on the hardware we had available. We
now report simulation results instead, as suggested.

We thank the reviewer for pointing out errors in the original
figures. These arose during manual conversion of indiegram's output
parse trees into Stockholm-format alignments, and due to a bug in
xrate's reconstruction code. We have now automated this process
(i.e. indiegram now outputs Stockholm format) and fixed the bug in
xrate, verifying that this has eliminated the errors (e.g. see the
simulation results).

We removed the tRNA and nanos reconstructions and posted (corrected)
versions online, at the following URL (which is linked from the
indiegram webpage):

http://biowiki.org/IndiegramExamples

We hope that these may still be useful for illustrative purposes for
anyone trying to use the software, but we have removed references from
the paper (since, as the reviewer notes, the short branch lengths make
the answers somewhat obvious).

The paper still includes the (experimentally-based) runtime estimates for
different length classes of RNA genes.

> More minor comments follow.
>
> title: A paper should deliver something akin to what its title seems 
> to promise. There is no result in this paper that bears meaningfully 
> on "primordial ribozymes" or "ancient RNA". The main results are 
> instead a promising but rough and very preliminary theoretical 
> framework for modeling the evolution of structural RNAs, with some 
> small toy examples of applications to three small RNA sequences. The 
> grand idea of paleogenetics motivates the work, and certainly should 
> be discussed, but it doesn't belong in the title unless you're 
> actually doing (and testing!) paleogenetics. A less grand title, more 
> consistent with results on toy three-sequence data sets, is appropriate.

We have accordingly changed the title to "Evolutionary models of
structured RNA".

> p.3: as far as I'm aware, the Felsenstein "pruning algorithm" and 
> "peeling algorithm" are just alternative names for the same algorithm 
> for recursively computing the likelihood of observed sequences related 
> by a tree, so I don't understand the distinctions being drawn in the 
> introduction between using the "pruning algorithm" to solve the 
> likelihood (problem "A") versus the "peeling algorithm" to sample from 
> the posterior (problem "C"). All four problems A-D are solved by the 
> same algorithm (or trivial postprocessing of its results) in the 
> ungapped, substitution-only case. See for example Felsenstein's book 
> "Inferring Phylogenies", p. 253, where he states the equivalence 
> between his pruning algorithm and Elston and Stewart's peeling algorithm.

We certainly agree that the peeling and pruning algorithms are
equivalent for the case of reversible independent-sites models, in the
following sense.  For a reversible model, the posterior distribution
of the state at any ancestral node can be obtained by re-rooting the
tree at that node and then performing the pruning algorithm,
traversing the tree post-order (terminating at the chosen root).
However, to then compute the posterior distribution of the state at a
*different* ancestral node would require re-rooting the tree at that
node and recomputing the entire DP matrix (or at least the part
corresponding to the branches whose direction had been reversed by the
re-rooting operation).  In contrast, the peeling algorithm yields all
such posterior probability distributions in a single traversal, so it
is more efficient (and distinct from pruning).

More generally, the two algorithms are distinct for the case of
irreversible models. More precisely, pruning is a restricted case of
peeling which only gives probabilities at the root node; since
irreversible models do not permit re-rooting of the tree, pruning
cannot (by itself) be used to provide posterior distributions for
internal nodes (only the root node).

We drew the distinction primarily to illustrate the different
applications for which the algorithms are frequently used (e.g., "A"
vs. "C"), but we can certainly revise this text further if the
reviewer feels that it is appropriate to do so.

> p.6 bottom: pair HMMs are not really "standard"; they are not widely 
> used or known. (There is still too much of a general tendency in the 
> manuscript to use technically narrow jargon in place of actual 
> explanation, and assume the reader will be able to fill in the gaps. 
> Sometimes, though, the revision gets it right: for example, other than 
> this "standard pair HMM" reference, the authors do an excellent job on 
> pp6-7 introducing the concept and motivation of conditional 
> probability models first, and *then* connecting to the linguistics 
> literature and the jargon term "parse-tree transducer".)

Thank you for pointing out this oversight.  We briefly describe what a
Pair HMM is in the revised manuscript.

> p.7 why is the model that generates P(X) at the root a "singlet 
> transducer"? It's not transducing anything (it's not conditionally 
> dependent on an input parse tree).

We chose to call it a "singlet" transducer so that we used similar
terminology for the model generating the marginal and conditional
distributions.  A singlet transducer can be viewed as a special case
of a transducer, where the input being transduced is empty.  Indeed,
by "singlet transducer" we mean the same thing as is often referred to
as a "single-sequence SCFG".  We apologize if it is poor terminology;
we can change it if the reviewer feels that doing so would help
clarify the methods.

> p.13 The DP inference algorithm section says "the standard CYK ... 
> algorithms fail". But then it says "Maximum likelihood inference is 
> performed with a CYK algorithm", and even "maximum likelihood 
> inference can therefore be performed with a standard three sequence 
> CYK algorithm". Which is it?

Thank you for pointing out this inconsistency, which we have clarified
in the revised text.  Using the standard CYK algorithm on the composed
model fails; however, the model can be simply modified as described in
the revised text to permit standard CYK to work properly.

> p.14 It seems possible that the problem with "null cycles" (residues 
> in ancestors that leave no homologous residues in observed sequences) 
> may simply go away if you view the problem as a missing data problem, 
> and that the likelihood being inferred is the likelihood of alignment 
> columns needed to account for the observed data (marginalized over 
> missing columns needed to account for null cycles) rather than the 
> joint likelihood of the "observed" columns and the "null cycle" 
> columns. This is what dropped out of the equations in the phylogenetic 
> inference described by Rivas and Eddy (2008) PLoS CB 4:e1000172, for 
> example.

We completely agree and have included an extended discussion of this in
the revised text.  The primary difficulty is that while this marginalization
is reasonably straightforward for independent-sites models and HMMs,
the possible presence of null bifurcations complicates matters for SCFGs.

> p.15-18 The whole section on "terminological asides" contributes 
> little to the paper. It would be vastly preferable to spend this space 
> describing what you're actually doing, as opposed to making vague 
> conceptual connections. Save the conceptual connections for a review 
> article someday.

We have accordingly stricken this section from the paper.

> p.19 bottom: "The results of these results, reported in Text S1...": 
> again, results in the paper should not be relegated to supplementary 
> material. A test that we use here is that you should be able to make 
> no references at all to Supplementary Material in the text of a paper, 
> but just list it at the end, and the reader would still be happy. Good 
> is "oh look, the source code and computer-readable data files are 
> available too", and bad is "dammit, you mean I have to go to another 
> manuscript to see the result of the experiment they just described?"

### 2009-04-18 response-to-reviews2.txt

Thank you for this suggestion.  We have moved those results back into
the main text of the paper and made them into a section discussing
evaluating the TKFST model by making it part of a multiple RNA
alignment program.

> p.21 "A TKFST-based reconstruction could in principle falsify 
> Hopfield's hypothesis": how, exactly? How could you falsify a 
> biological hypothesis with a model whose realism and accuracy you've 
> called into question?

[we have deleted the references to Hopfield's hypothesis]

> p.21 in general the whole "reconstructing RNAs with a simple model" 
> section is too chatty and imprecise. "Both predicted ancestral 
> structures closely match the RFAM-annotated consensus structures", for 
> example, is not a quantitative result, and was not set up to address 
> any particular focused question (like, for instance, does our method 
> do better than some simpler approach of addressing the same problem?) 
> Again too much space is devoted to vague forward-looking conceptual 
> claims ("Though it's beyond the scope of the paper, a revised pairwise 
> model could readily be designed...") and too little attention is given 
> to the actual model and problems at hand.

> Table 2: it's unclear why we need separate states L and I_L, or S and 
> I_S; these nonterminals have identical production rules.

[distinction between clearest representation of the grammar for didactic purposes, versus the best representation for implementation purposes; for implementation we used a Moore machine representation]

> Similarly, I 
> don't see why we need B_e; S and I_S can produce L directly; and L and 
> I_L can produce an LS bifurcation directly, rather than invoking 
> nonterminal B. The L and I_L productions don't appear to be correctly 
> normalized, unless the (undefined) term p_S is 0.0.

> A production I_L 
> \rightarrow x I_L does not correspond to a pair-HMM rule, as the 
> caption states; this is just a standard regular grammar production.

Thank you for pointing this out.  We have corrected this in the 
revised text.

> Table 4: perhaps I'm slow, but this doesn't seem like the clearest 
> explanation of a branch transducer. I'm expecting to see rules 
> corresponding to, for example, the ancestor produced G-C from a 
> basepair state; the branch transducer chooses a basepair->basepair 
> substitution transduction as its state path with some probability, 
> then eats the GC and emits an xy. The production rules in Table 4 look 
> more like a single-sequence SCFG to me (x,y do appear in the wait 
> state conditionals, but where are x,y coming from on the left hand 
> side -- where are they being absorbed from?)

> The productions also 
> don't seem to correspond entirely to the state definitions in Table 5; 
> for example, if M_S is a "match" state describing "base pair 
> substitution" that "absorbs (x,y)" and "emits (u,v)" according to 
> Table 5, then why does the Table 4 give no production rule for 
> emitting (u,v) given (x,y)? Instead it's the wait state W_S in Table 4 
> that has the expected (x,y) -> (u,v) production rule. Maybe the
> rationale would become clear to me if I spend more time staring at 
> this, but it seems to need more explanation and/or clarification and 
> cleanup.

We wrote the rules here in the Mealy-machine framework, where inputs
and outputs are attached to transitions (c.f., the Moore-machine framework,
where ouputs depend only on the current state of the machine).  The
rule
W_S -> u M_s v ;       alpha_2 (t) * p (u, v| x, y)
is intended to correspond to a transition in a Mealy machine
from state W_S to M_S, where the transition has an input of (x, y)
and an output of (u, v).  We have attempted to clarify this in the
revised text.

We apologize for the confusion; we have attempted to follow the
standard SCFG notation as closely as possible in an attempt to be
clear, but we certainly welcome further suggestions to clarify
the model.

> Figure 1: 5' and 3' ends of RNA structures should be labeled.

> Figure 3: It is very unclear what this figure is supposed to be 
> representing (and I had the same problem with it when it appeared in 
> reference [28]). It appears to simply be a representation of a parse 
> tree of an individual RNA structure. It does not seem to show any kind 
> of evolutionary model from one sequence/structure/parse tree to 
> another, as the caption and text would seem to imply.

[we have included stree-mutations.pdf, which was also in the cited paper, and illustrates the dynamic behavior of this model]

> Figure 4 and Figure 5 seem unnecessary. Both are too schematic to 
> convey useful information. The concept of conditional likelihood along 
> branches of a tree is obvious and well understood to your readers in 
> the substitution-only framework of the Felsenstein algorithm. The 
> authors should focus these figures like this on their *particular* 
> model for the transducers, and exactly how they accomodate insertion 
> and deletion, rather than the obvious concept of "P(X | Y,t)".

We felt that it was better to include information at the risk of
stating the obvious, rather than risk omitting necessary
information. The figure illustrating "P(X | Y,t)" may be obvious but
the figure illustrating the following distribution...

P(W | root_distribution) P(X | W,T_X) P(Y | W,T_Y) P(Z | W,T_Z)

...is, we feel, not as obvious; and its simplicity may help clarify
the more complicated concepts about transducer composition that appear
in the text. Our hope is that since this is listed as a minor
criticism, we may accept it as a stylistic difference.

> The figures are referred to out of order; I try not to succumb to the 
> hobgoblins of foolish consistencies, but it's pretty egregious here (I 
> think Figure 1 isn't referred to until p.21; whereas I think Figure 4 
> is mentioned first, on p.9.) Similarly, the bibliography seems 
> excessively typo-laden. (Ref 25 lists the author as "J P". Ref 28 
> gives no page numbers. Ref 33 shows no volume nor page numbers. And so 
> on.)

[we have corrected this..... I hope]
